---
title: "Why factor analysis is a poor test of unidimensionality"
author: "Johannes Titz"
date: "2/23/2022"
output: 
  pdf_document: 
    fig_caption: yes
    latex_engine: lualatex
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE,
                      dev = c("cairo_pdf", "png"),
                      fig.path = "plots/tm",
                      dev.args = list(antialias = "subpixel"),
                      dpi = 300,
                      echo = TRUE,
                      fig.height = 4.365, fig.width = 4.85,
                      options(width = 82))
```

# Prepare

```{r prepare}
library(librarian)
shelf(partitions, pbapply, tidyverse, psych, johannes-titz/zysno, xtable)
# use multiple cores
cores <- round(parallel::detectCores() * 0.8)
zys <- function(a, b, c, d) {
  n <- a + b + c + d
  e <- 2 * b * c
  exp_e <- 0.5 * n^2 * (1 - ((a + c)^2 + (b + d)^2) / n^2) *
      (1 - ((a + b)^2 + (c + d)^2) / n^2)
  1 - e / exp_e
}
get_h <- function(c, n, r, a, d) {
  1 - c^2 * n^2 * r^2 / (a * d - c^2)^2
}
# create vectors from phi matrix
vec_from_tbl <- function(a, b, c, d) {
  mtrx <- matrix(c(a, b, c, d), ncol = 2)
  v1 <- rep(c(0, 1), colSums(mtrx))
  v2 <- rep(c(0, 1), rowSums(mtrx))
  cbind(v1, v2)
}
```

## Binary Variables

### Generate Data

```{r generate, cache = F}
n <- 50
# all combos, even the ones where phi is NA
grid <- as.matrix(t(compositions(n = n, m = 4)))
colnames(grid) <- paste0("Var", 1:4)
grid <- as_tibble(grid) %>%
  rowwise() %>%
  mutate(tetrachora = cos(pi * sqrt(Var2 * Var3) / (sqrt(Var1 * Var4) + sqrt(Var2 * Var3))),
         errors = Var3,
         expected_errors = (Var1 + Var3) * (Var3 + Var4) / n,
         h = 1 - errors / expected_errors,
         hz = zys(Var1, Var2, Var3, Var4))
# rowwise does not work with the following code, I do not know why
# multiple cores seem to give no advantage, so is not used (argument cl)
# todo: suppressWarnings does not seem to work
# cache it
grid <- xfun::cache_rds({
  phi <- pbapply(grid[, 1:4], 1, function(x) phi(x, digits = 4))
  tetrachor <- pbapply(
    grid[, 1:4],
    1,
    function(x) tetrachoric(x, correct = 0)$rho
  )
  # do not display warnings for corrections
  tetrachor_corrected <- pbapply(
    grid[, 1:4],
    1,
    function(x) suppressWarnings(tetrachoric(x, correct = 0.5)$rho)
  )
  cbind(grid, phi, tetrachor, tetrachor_corrected)
}, file = "sim1")
grid <- as_tibble(grid)
```

```{r filter}
grid_unidim <-  grid %>%
  filter(Var3 == 0) # only unidimensional
grid_row <- grid %>%
  filter(phi >= 0, Var2 >= Var3) # error cell is b
```

### Plots 

These two plots should be enough for a publication:

```{r phihs}
ggplot(grid_row, aes(phi, h)) + geom_point(alpha = 0.2) +
  theme_classic()
```

The same, but for tetrachoric and tetrachoric corrected.

```{r tetrah}
grid_tet <- grid %>%
  filter(tetrachor >= 0, Var2 >= Var3)
grid_tetc <- grid %>%
  filter(tetrachor_corrected >= 0, Var2 >= Var3)
ggplot(grid_tet, aes(tetrachor, h)) + geom_point(alpha = 0.2) +
  theme_classic()
ggplot(grid_tetc, aes(tetrachor_corrected, h)) + geom_point(alpha = 0.2) +
  theme_classic()
```

### FA model

Now, it is not hard to create unidimensional model that results in a perfect bifactor solution with factor analysis (for Pearson Correlation):

```{r}
m1 <- vec_from_tbl(10, 80, 0, 10)
m2 <- vec_from_tbl(11, 78, 0, 11)
m3 <- vec_from_tbl(15, 70, 0, 15)

d <- cbind(m1, m2, m3[, -2])
ncol <- ncol(d)
colnames(d) <- paste("Item", seq(ncol(d)))
f1 <- factanal(d, 1)
f2 <- factanal(d, 2)
print(xtable::xtable(cbind(f1$loadings, f2$loadings)),
      file = "tables/fa1b.tex", booktabs = TRUE)
combs <- Map(function(x, y) table(d[, x], d[, y]),
             rep(1:ncol, each = ncol), rep(1:ncol, ncol))
lower <- seq(1, 21, 5)
upper <- seq(5, 25, 5)
res <- Map(function(x, y) Reduce(cbind, combs[x:y]),
    lower, upper)
res2 <- Reduce(rbind, res)
colnames(res2) <- paste0("$", rep(1:5, each = 2), "_", c(0, 1), "$")
rownames(res2) <- paste0("$", rep(1:5, each = 2), "_", c(0, 1), "$")
print(xtable::xtable(res2), booktabs = TRUE,
      file = "tables/fa1.tex",
      sanitize.text.function = function(x) {x})
patterns <- table(apply(d, 1, paste, collapse = ""))
```

## three item categories 

Problem with H value is that we need to find the error cells first, this takes some additional time. But seems to be ok overall.

```{r poly, cache = F}
shelf(partitions)
test <- xfun::cache_rds(
  {t(compositions(n = 10, m = 16))},
  file = "partitions"
)
dim(test)
create_d <- function(f) {
  cols <- colSums(matrix(f, nrow = 4))
  rows <- rowSums(matrix(f, nrow = 4))
  v1 <- rep(1:4, cols)
  v2 <- rep(rep(1:4, 4), f)
  return(cbind(v1, v2))
}
get_values <- function(d) {
  h <- zysno:::lv_errors_item_pair(d)$h
  cor <- cor(d)[1, 2]
  # 0 variance
  if (!is.na(cor)) {
        tetrachor <- polychoric(d, correct = FALSE)$rho[1, 2]
        tetrachor_corrected <- tryCatch(
          {polychoric(d)$rho[1, 2]},
          warning = function(cond) {
            #message(warning)
            return(NA)
          },
          error = function(cond) {
            #message(cond)
            return(NA)
          })
  } else {
    tetrachor <- NA
    tetrachor_corrected <- NA
  }
  return(c(cor, h, tetrachor, tetrachor_corrected))
}
res <- xfun::cache_rds({
  pblapply(seq(nrow(test)), function(x) get_values(create_d(test[x, ])), cl = 29)
}, file = "sim2", rerun = T)
resm <- matrix(as.numeric(unlist(res)), ncol = 4, byrow = TRUE)
resm <- apply(resm, 2, round, 2)
df <- as.data.frame(resm)
colnames(df) <- c("cor", "h", "tetrachor")

# otherwise plotting is impossible
df1 <- df[!duplicated(df[, 1:2]), ]
df2 <- df[!duplicated(df[, c(1:3)]), ]
df1 <- df1[df1[, 1] >= 0, ]
df2 <- df2[df2[, 3] >= 0, ]
names(df1) <- c("cor", "h", "tetrachor")
names(df2) <- c("cor", "h", "tetrachor")
plot(df1[, 1], df1[, 2])
plot(df2[, 3], df2[, 2])
```

```{r}
ggplot(df1, aes(cor, h, alpha = 0.1)) + geom_point() + theme_classic()
ggplot(df2, aes(tetrachor, h, alpha = 0.1)) + geom_point() + theme_classic()
```

## FA model

Seems to be a problem with calculation that needs to be fixed first.
```{r}
filter <- which(df$h == 1 & df$tetrachor <= 0.1 & df$tetrachor > 0)
df[filter, ]
part <- test[filter, ]
part2 <- t(apply(part, 1, function(x) ifelse(x != 0, 1, 0)))
filter2 <- which.max(rowSums(part2))
part[filter2,]
dfa <- create_d(part[filter2,])
cor(dfa)
polychoric(dfa, correct = F)
```
