---
title: "Why factor analysis is a poor test of unidimensionality"
author: "Johannes Titz"
date: "2/23/2022"
output: 
  pdf_document: 
    fig_caption: yes
    latex_engine: lualatex
    keep_tex: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, cache = TRUE,
                      dev = c("cairo_pdf", "png"),
                      fig.path = "plots/tm",
                      dev.args = list(antialias = "subpixel"),
                      dpi = 300,
                      echo = TRUE,
                      fig.height = 4.365, fig.width = 4.85,
                      options(width = 82))
```

# Prepare

```{r prepare}
library(partitions)
library(tidyverse)
library(psych)
zys <- function(a, b, c, d) {
  n <- a + b + c + d
  e <- 2 * b * c
  exp_e <- 0.5 * n^2 * (1 - ((a + c)^2 + (b + d)^2) / n^2) *
      (1 - ((a + b)^2 + (c + d)^2) / n^2)
  1 - e / exp_e
}
get_h <- function(c, n, r, a, d) {
  1 - c^2 * n^2 * r^2 / (a * d - c^2)^2
}
# create vectors from confusion matrix
vec_from_tbl <- function(a, b, c, d) {
  mtrx <- matrix(c(a, b, c, d), ncol = 2)
  v1 <- rep(c(0, 1), colSums(mtrx))
  v2 <- rep(c(0, 1), rowSums(mtrx))
  cbind(v1, v2)
}
```

## Binary Variables

### Generate Data

```{r generate}
n <- 50
# all combos, even the ones where phi is NA
grid <- as.matrix(t(compositions(n = n, m = 4)))
colnames(grid) <- paste0("Var", 1:4)
# for case 1
grid <- as_tibble(grid) %>%
  rowwise() %>%
  mutate(tetrachor = tetrachoric(c(Var1, Var2, Var3, Var4), correct=0)$rho,
         tetrachor_corrected = tetrachoric(c(Var1, Var2, Var3, Var4))$rho,
tetrachora = cos(pi*sqrt(Var2*Var3) / (sqrt(Var1*Var4) + sqrt(Var2*Var3))),
phi = phi(c(Var1, Var2, Var3, Var4), digits = 5),
         errors = Var3,
         expected_errors = (Var1 + Var3) * (Var3 + Var4) / n,
         h = 1 - errors / expected_errors,
         hz = zys(Var1, Var2, Var3, Var4))
```

```{r filter}
grid2 <-  grid %>%
  filter(Var3 == 0) # only unidimensional
grid3 <- grid %>%
  filter(phi >= 0, Var2 >= Var3) # error cell is b 
```

### Plots

This plot should be enough for a publication:

```{r phihs}
ggplot(grid3, aes(phi, h)) + geom_point(alpha=0.2) +
  theme_classic()
```

Some additional plots:

As long as the sample size is not restricted, every correlation is possible. For $N=100$:

todo: add explanation with arrow to one b, c=0

```{r, unidimphi}
pos_c <- grid2 %>%
  filter(!is.na(phi)) %>%
  group_by(Var1, Var2) %>%
  summarize(min_phi = min(phi, na.rm = T)) %>%
  slice(which.min(min_phi))

p1 <- ggplot(grid2, aes(Var1, phi, color = Var4, group = Var2)) +
  geom_point(size = 1) + theme_classic() + geom_line() +
  scale_color_continuous(name = "d", low = "darkgray", high = "black") + 
  scale_x_continuous(name="a")

p1 + geom_text(data=pos_c, mapping=aes(Var1-0.01, min_phi-0.015, color=NULL, label = Var2))
```

### tetrachoric correlation

```{r unidimtetra}
pos_c <- grid2 %>% 
  filter(!is.na(tetrachor)) %>%
  group_by(Var1, Var2) %>%
  summarize(min_phi = min(tetrachor, na.rm=T)) %>%
  slice(which.min(min_phi))

ggplot(grid2, aes(Var1, tetrachor, col = Var3)) + 
  geom_point() + theme_classic()

p1 <- ggplot(grid2, aes(Var1, tetrachor, color = Var4, group = Var2)) + 
  geom_point(size = 1) + theme_classic() + geom_line() +
  scale_color_continuous(name = "d", low = "darkgray", high = "black") + 
  scale_x_continuous(name="a")

p1 + geom_text(data=pos_c, mapping=aes(Var1-0.02, min_phi-0.03, color=NULL, label = Var2))
```

### FA model

Now, it is not hard to create unidimensional model that results in a perfect bifactor solution with factor analysis: 

```{r}
m1 <- vec_from_tbl(10, 80, 0, 10)
m2 <- vec_from_tbl(11, 78, 0, 11)
m3 <- vec_from_tbl(15, 70, 0, 15)

d <- cbind(m1, m2, m3[, -2])
ncol <- ncol(d)
colnames(d) <- paste("Item", seq(ncol(d)))
f1 <- factanal(d, 1)
f2 <- factanal(d, 2)
print(xtable::xtable(cbind(f1$loadings, f2$loadings)), file = "tables/fa1b.tex", booktabs = T)
combs <- Map(function(x, y) table(d[, x], d[, y]), rep(1:ncol, each = ncol), rep(1:ncol, ncol))
lower <- seq(1, 21, 5)
upper <- seq(5, 25, 5)
res <- Map(function(x, y) Reduce(cbind, combs[x:y]),
    lower, upper)
res2 <- Reduce(rbind, res)
colnames(res2) <- paste0("$", rep(1:5, each = 2), "_", c(0, 1), "$")
rownames(res2) <- paste0("$", rep(1:5, each = 2), "_", c(0, 1), "$")
print(xtable::xtable(res2), booktabs = T, file = "tables/fa1.tex",
sanitize.text.function=function(x){x})
patterns <- table(apply(d, 1, paste, collapse = ""))
```

## second simulation

```{r}
ggplot(grid3, aes(phi, errors)) + geom_point(alpha=0.2) +
  theme_classic()
```


```{r phihsupper}
grid3b <- grid3 %>% filter(Var3 == 1, Var1 == Var4)
ggplot(grid3b, aes(phi, h)) + geom_point(alpha=0.2) +
  theme_classic()
```

```{r phihz}
ggplot(grid3, aes(phi, hz)) + geom_point(alpha = 0.2) +
  theme_classic()
```

```{r tetrah}
grid_tet <- grid %>%
  filter(tetrachor >= 0, Var2 >= Var3)
grid_tetc <- grid %>%
  filter(tetrachor_corrected >= 0, Var2 >= Var3)
ggplot(grid_tet, aes(tetrachor, h)) + geom_point(alpha = 0.2) +
  theme_classic()
ggplot(grid_tetc, aes(tetrachor_corrected, h)) + geom_point(alpha = 0.2) +
  theme_classic()
```

```{r tetraha}
grid_tet <- grid %>% filter(tetrachora >=0, Var2 >= Var3)
ggplot(grid_tet, aes(tetrachora, h)) + geom_point(alpha=0.2) +
  theme_classic() 
```

```{r tetrahz}
ggplot(grid_tet, aes(tetrachor, hz)) + geom_point(alpha=0.2) +
  theme_classic() 
```

##  relationship errors and phi

```{r}
maxe <- grid3 %>% group_by(errors) %>% summarize(max_phi = max(phi))
plot(maxe$max_phi, maxe$errors)
m1 <-  lm(errors/n~max_phi, data = maxe)
summary(m1)
```

Wenn r=0 u. r =1
e/N = b0+b1*r
0 = b0+b1*1
1/4 = b0+0

Wenn r=1, dann alles auf diagonale ad, wenn r=0, alles gleich verteilt. es gilt bc <= ad (nur positive korrelationen!). es gilt b>=c (in der Zeile ist das einfachere Item).

# three item categories 

todo: run on server to get more values?

```{r poly}
librarian::shelf(partitions)
test <- t(compositions(n = 12, m = 9))
dim(test)
create_d <- function(f){
  cols <- colSums(matrix(f, nrow = 3))
  rows <- rowSums(matrix(f, nrow = 3))
  # every item category should have at least one value
  # debatable
#  if (any(cols==0) | any(rows==0)) return(NA)
#  if (sum(f > 0) <2) return(NA)
  v1 <- rep(1:3, cols)
  v2 <- rep(rep(1:3, 3), f)
  return(cbind(v1, v2))
}

# ugly with for loop 
l <- list()
for (i in seq(nrow(test))) {
  d  <- create_d(test[i,])
  l[[i]] <- d
}

l2 <- l[!is.na(l)] # omit NA

get_hz <- function(d) {
  act <- sum(zysno:::get_errors_for_one_item_pair(d))
  exp <- zysno:::get_expected_errors(d)
  cor <- cor(d)[1,2]
  if (!is.na(cor)) {
        tetrachor <- polychoric(d, correct = F)$rho[1,2] # takes forever
} else tetrachor <- NA
  return(data.frame(cor=cor, hz = 1-act/exp, act = act, tetrachor = tetrachor))
}
librarian::shelf(pbapply)
hz <- pbapply::pblapply(l2, get_hz)
#hz <- parallel::mclapply(l2, get_hz)
df <- plyr::ldply(hz, "data.frame")
df <- df[df$cor > 0,]
plot(df$cor, df$hz)
plot(df$tetrachor, df$hz)
```

Homogenity index can become negative. should it be set to 0 then?

Create an example for FA, which is undimensional but produces two factors

```{r}
df[df$hz==1 & df$cor<0.25,]
table(as.data.frame(l2[[11858]]))
```

## relationship homogenity zsyno

when b==c, we get the minimal hz value (most errors)

```{r}
grid3b <- grid3 %>% 
  filter(Var2==Var3)#, Var2 == 1)
head(grid3b[,-5])
plot(grid3b$phi, grid3b$hz)
head(grid3b)
grid3b
```

# appendix

## dichotomisierung

Frage: können wir Korrelation und Fehler aus dichotomisierten Items konstruieren?

Anscheinend nicht, denn orthogonale Variablen müssen nicht mit 0 korrelieren? Sehr merkwürdig! Zwar ist dann im linearen Modell die Korrelation korrekt, aber aus den einzelnen Dummy-Variablen ist das nicht ersichtlich.

```{r}
v1 <- psych::dummy.code(zysno47[,1])
v2 <- psych::dummy.code(zysno47[,2])
cor(v1)
cor(zysno47) # 0.74
summary(lm(zysno47[,2]~v1[,1]+v1[,2]))
sqrt(0.5499) # 0.74
```

## some tries with 3d plots

```{r, eval=F}
librarian::shelf(plotly)
librarian::shelf(AckerDWM/gg3D)
librarian::shelf(rayshader)
options(browser="firefox")
set.seed(417)
library(plotly)
temp <- rnorm(100, mean=30, sd=5)
pressure <- rnorm(100)
dtime <- 1:100

plot_ly(grid2, x=~Var1, y=~Var4, z=~phi, mode="markers", color=~Var2,
        size = 1)#  %>% group_by(Var4) %>% add_lines()

ggplot(grid2, aes(x=Var1, y = Var4, z = phi, colour = Var2))+
  axes_3D() +
  stat_3D()

mtplot = ggplot(mtcars, aes(x = mpg, y = disp, group = gear, color = cyl)) + 
  geom_point() + geom_line()+
  scale_color_continuous(limits = c(0, 8))
mtplot
#par(mfrow = c(1, 2))
#plot_gg(mtplot, width = 3.5, raytrace = FALSE, preview = TRUE)

plot_gg(mtplot, multicore = TRUE)#, width = 3.5, multicore = TRUE, 
        zoom = 0.85, phi = 35, theta = 30, sunangle = 225, soliddepth = -100)
Sys.sleep(0.2)
render_snapshot(clear = TRUE)
```
